%----------------------------------------------------------------------------------------
%	BOOK INFORMATION
%----------------------------------------------------------------------------------------

% \titlehead{Document Template}
\title[CN:\ Theory and Engineering for Verifying C]{CN:\ Theory and Engineering for Verifying C}
% \subtitle{}
\author[Dhruv Makwana]{Dhruv Makwana}%
\date{August 2026}
\publishers{This dissertation is submitted for the degree of Doctor of Philosophy}

%----------------------------------------------------------------------------------------

\frontmatter % Denotes the start of the pre-document content, uses roman numerals
\pagenumbering{arabic}

%----------------------------------------------------------------------------------------
%	COPYRIGHT PAGE
%----------------------------------------------------------------------------------------

% \makeatletter
% \uppertitleback{\@titlehead} % Header
%
% \lowertitleback{%
% 	% \textbf{Disclaimer} \\
% 	% You can edit this page to suit your needs. For instance, here we have a no copyright statement, a colophon and some other information. This page is based on the corresponding page of Ken Arroyo Ohori's thesis, with minimal changes.
%
% 	% \medskip
%
% 	% \textbf{No copyright} \\
% 	% \cczero\ This book is released into the public domain using the CC0 code. To the extent possible under law, I waive all copyright and related or neighbouring rights to this work.
%
% 	% To view a copy of the CC0 code, visit: \\\url{http://creativecommons.org/publicdomain/zero/1.0/}
%
% 	% \medskip
%
% 	% \textbf{Colophon} \\
% 	% This document was typeset with the help of \href{https://sourceforge.net/projects/koma-script/}{\KOMAScript} and \href{https://www.latex-project.org/}{\LaTeX} using the \href{https://github.com/fmarotta/kaobook/}{kaobook} class.
%
% 	% \medskip
%
% 	% \textbf{Publisher} \\
% 	% First printed in May 2019 by \@publishers
% }
% \makeatother

%----------------------------------------------------------------------------------------
%	DEDICATION
%----------------------------------------------------------------------------------------

\dedication{\raggedleft\emph{To all those who supported and believed in me.}}

% \dedication{
% 	% The harmony of the world is made manifest in Form and Number, and the heart and soul and all the poetry of Natural Philosophy are embodied in the concept of mathematical beauty.\\
% 	% \flushright -- D'Arcy Wentworth Thompson
% }

%----------------------------------------------------------------------------------------
%	OUTPUT TITLE PAGE AND PREVIOUS
%----------------------------------------------------------------------------------------

% \includepdf[pages=1]{../MathSTICTemplate/main.pdf}

% Note that \maketitle outputs the pages before here
\maketitle

%----------------------------------------------------------------------------------------
%	PREFACE
%----------------------------------------------------------------------------------------

\begingroup

  % in the header, chapters start on any page to avoid too many blank pages
  \let\cleardoublepage\clearpage

\addchap{Declaration}

This thesis is the result of my own work and includes nothing which is the
outcome of work done in collaboration except as declared in the Preface and
specified in the text. I further state that no substantial part of my thesis
has already been submitted, or, is being concurrently submitted for any such
degree, diploma or other qualification at the University of Cambridge or any
other University or similar institution except as declared in the Preface and
specified in the text.

It does not exceed 66,000 words including tables, footnotes and equations, but
excluding appendices, bibliography, photographs and diagrams.

Previous publications and collaborations are detailed in the Introduction.

\addchap{Abstract}

More than fifty years after its introduction, \kl{C} remains in common use,
especially for low-level software such as \kl{operating system} \kl{kernel}s.
\kl{Kernel}s need to be both \emph{fast} and \emph{secure}, thus need to
be written in languages which expose lots of control. However, more human
control means more chance for human error. Historically, both hardware and
software were much rarer, simpler and more trusted, but these days errors pose
a multi-million computer, multi-billion dollar risk.

The advent of separation logic enabled the pen-and-paper verification of
low-level programs in an idealised imperative language; its mechanisation in
proof tools enabled machine-checked proofs of the same. With the advent of
\kl{Cerberus}, an executable and empirically validated semantics for C, we now
have the opportunity to drop the idealisation

In this thesis, I show how to build a verification tool for C, called \kl{CN},
suitable for handling low-level systems programming idioms, and argue that
doing so requires some theory and lots of engineering.

In the first part, I describe \kl{CN}'s design goals and constraints, and how
these informed the theory required. I discuss the challenges I faced in
applying that theory to the \kl{Cerberus} semantics,  and how I solved them. I
also discuss how the theory fed back insights to engineering questions which
arose during the development of \kl{CN}.

In the second part, I describe how to handle the \kl{memory object model} for
C, which is necessary to match existing compiler behaviour, and support
low-level programming idioms. I explain the difficulty in needing to balance
simplicity and usability of \kl{CN} with expressiveness. Here too, the theory
fed back insights on how to improve performance and add new features.

In the last part, I describe how to tame some of the complexity of large
repositories and non-standard features of real-world C. Then, I explain why,
despite being on track to make proofs of low-level C code \emph{feasible},
\kl{CN} still has some way to go before to become more \emph{usable}, and its
proofs more maintainable. After that, I compare \kl{CN} to other C verification
tools, and discuss feedback on \kl{CN} by partners in industry. I conclude by
sketching out longer-term lines of research on such tools.

With this work, I show that building a verification tool for C, which can
handle low-level systems programming idioms is tractable, and that there are
still plenty of avenues for substantial improvement.

\addchap{Acknowledgements}

These past six years have reoriented my life dramatically, academically,
professionally and personally. This would not have happened without the support
of many people in my life.

First I thank my supervisor, Neel Krishnaswami, for his support from
essentially the day he arrived at the University of Cambridge. He nurtured and
directed my interest in linear types while I was a mere undergraduate, and has
supported me in ways too numerous to list here since. I would also like to
thank my co-supervisor, Peter Sewell, for his hands-on guidance and high
standards. I am extra grateful to both for supporting my impromptu move to
Wellington, and multiple long meditation retreats once I came back.

\k{CN} has been collaborative from the start, and so I must also thank my
colleagues, Christopher, Thomas Sewell, Kayvan, Rini, Hiro, and everyone else
in the REMS group, especially Nils, who extended their friendship and kindness,
livened up the weekly meetings, and shared my love of Indian mangoes. I also
thank my office mates for their company and life-lessons.

Next, I want to thank all those who made my time in Wellington, New Zealand, at
the start of my PhD, so dear and memorable. All of you helped transform what
could have been the worst time in my life to the best: my housemates Dylan,
Dan, and Laura, my friends from the animal activist, Bhakti Lounge, effective
Altruism, Bodhinyanarama, James Noble, my colleagues at the most scenic
graduate building in the world, and Latin dance communities. I miss you all
frequently.

I similarly wish to thank my friends in the rest of the world for the ongoing
support and variety they brought to my life: from school, undergrad, Bollywood
Dance Troupe, Plant-based Cambridge, local animal activism groups, Aurelius
Foundation, Effective Altruism, Buddhist Society, Cambridge Interbeing Sangha,
Amaravati, and all my meditation retreats.

Lastly, and most importantly, I wish to thank my family: my parents, my sister,
my cousins, nieces, nephew, uncles, aunts and grandparents. The older I get the
more I appreciate what it means to have people who are happy to see you and
ready to welcome you.

\pagelayout{margin}%
\addchap{How to Read This Thesis}

I anticipate that most of my readers will read this thesis on a device rather
than in print. As such, I make frequent use of hyperlinking, though without
visual cues to keep the text readable. For example, the word `bidirectional'
links to the place in~\fancyref{chap:formal-background} where it is defined,
and the definition itself is emphasised like this \intro{example}.

I also use large margins in the main text for notes, small figures, and
references. This is to reduce the need to teleport between the main text and
information that is far away.\sidenote[][0pt]{It is also is useful for including
supplementary information, whilst not breaking the flow of the text.}
For medium and large figures, I have done my best to keep them as near as
possible to their explanations. To achieve this, I have occasionally split a
figure into multiple parts, such as in
\cref{fig:minicn-return,fig:minicn-bop,fig:minicn-if}. In such instances, you
can interpret the figures as different parts of a whole, one rule or
non-terminal at a time.

Lastly, though I use colour to ease readability, for accessibility (and
printing), it should not be used to convey information.

\endgroup

%----------------------------------------------------------------------------------------
%	TABLE OF CONTENTS & LIST OF FIGURES/TABLES
%----------------------------------------------------------------------------------------

\pagelayout{wide}
\cleardoubleevenemptypage{}

\knowledgeconfigure{protect links}

\begingroup % Local scope for the following commands

\hypersetup{allcolors=.}

% Define the style for the TOC, LOF, and LOT
%\setstretch{1} % Uncomment to modify line spacing in the ToC
%\hypersetup{linkcolor=blue} % Uncomment to set the colour of links in the ToC
\setlength{\textheight}{230\vscale} % Manually adjust the height of the ToC pages

% Turn on compatibility mode for the etoc package
\etocstandarddisplaystyle% "toc display" as if etoc was not loaded
\etocstandardlines% "toc lines as if etoc was not loaded
\setcounter{tocdepth}{\sectiontocdepth} % Locally for the global toc

\tableofcontents % Output the table of contents

\setcounter{tocdepth}{\subsectiontocdepth} % To have correct tocs in the sections

% \listoffigures % Output the list of figures

% Comment both of the following lines to have the LOF and the LOT on different pages
% \let\cleardoublepage\bigskip
% \let\clearpage\bigskip

% \listoftables % Output the list of tables

\endgroup
\knowledgeconfigure{unprotect links}
